Online {\IterComp}

### 2017-09-14 infra.tex
arrrrrrrrrrrrrrrrrrrrrrrrrrrrrrrgh
minor fix

\begin{figure}[htb]
    \centering
    \includegraphics[width=\linewidth]{figs/infra-diagram}
    \caption{Overview of the online \itercomp mechanism.}
    \label{fig:infra-diagram}
\end{figure}

@@

This work focuses on the highlighted components of Figure~\ref{fig:infra-diagram}. The first one, \textit{Work Instrumentation} allows
measuring the amount of work performed. Conceptually, we use the cost model of Section~\ref{sec:metric} to estimate a cost for each basic
block of the IR and then add instrumentation to increase a global work counter by that cost every time a basic block is executed. The
measured work should depend only on the program and the input, not the applied optimizations, so we instrument the unoptimized IR before
any optimizations modify it. We present a detailed description of work instrumentation and how to keep it lightweight in
Section~\ref{chap:instr}.

@@

\FIXME{WHAT IS THE OPTIMIZATION EVALUATOR? WHY IS IT HIGHLIGHTED? DO WE HAVE ANYTHING TO SAY ABOUT IT?}

@@

Most devices have distinct idle and peak usage periods. For example, mobile devices are used intensely during the day and not at all while
the user is sleeping and the battery is charging~\citep{mpeis16}. Many authors have identified similar idle and peak usage periods in the
context of data centers~\citep{armbrust10,chen12b}. The proposed \itercomp infrastructure is very well suited to such scenarios. In
particular, we can use periods of idleness or underutilization to select the next few optimization sequences to try and compile the
corresponding binaries. This approach effectively hides the compilation overhead from the user. During peak periods, we only have to select
which of the existing binaries to evaluate which introduces almost no overhead.

>>>>

\begin{figure}[t]
    \centering
    \includegraphics[width=\linewidth]{figs/infra-diagram}
    \caption{Overview of the online \itercomp mechanism. \red{zw: mark the boundary of online and offline.}}
    \label{fig:infra-diagram}
\end{figure}

@@

This work focuses on the highlighted components of Figure~\ref{fig:infra-diagram}. The first one, \textit{Work Instrumentation} \blue{is
performed once at offline compile time}. It allows measuring the amount of work performed. Conceptually, we use the cost model of
Section~\ref{sec:metric} to estimate a cost for each basic block of the IR and then add instrumentation to increase a global work counter
by that cost every time a basic block is executed. The measured work should depend only on the program and the input, not the applied
optimizations, so we instrument the unoptimized IR before any optimizations modify it. We present a detailed description of work
instrumentation and how to keep it lightweight in Section~\ref{chap:instr}.

@@

%\FIXME{WHAT IS THE OPTIMIZATION EVALUATOR? WHY IS IT HIGHLIGHTED? DO WE HAVE ANYTHING TO SAY ABOUT IT?}

@@

Most devices have distinct idle and peak usage periods. For example, mobile devices are used intensely during the day and not at all while
the user is sleeping and the battery is charging~\citep{mpeis16}. Many researchers have identified similar idle and peak usage periods in
the context of data centers~\citep{armbrust10,chen12b}. The proposed \itercomp infrastructure is very well suited to such scenarios. In
particular, we can use periods of idleness or underutilization to select the next few optimization sequences to try and compile the
corresponding binaries. This approach effectively hides the compilation overhead from the user. During peak periods, we only have to select
which of the existing binaries to evaluate which introduces almost no overhead.

### 2017-09-14 prof.tex
minor fix

In this section, we describe how to measure work online. Total work is just the sum of the costs of all executed basic blocks, so it is
possible to embed its computation into the execution of the program. A naive instrumentation would use a global counter initialized at the
interception value, $\varepsilon$. Each basic block would then have only to increment this counter with the total cost of its instructions
every time it is executed. Although easy to implement, the naive approach introduces a significant overhead.

Previous work on basic block profiling has proposed optimal ways for instrumenting the code with as little overhead as possible without
losing any profiling information~\citep{knuth73,ball94}. The underlying idea is to instrument only a subset of the basic blocks of a
function and use the function's CFG to calculate execution counts for the rest. We can use the same idea for our purposes. Work profiling
and basic block profiling are similar, both need to count how many times each basic block was executed. The only important difference is
that we do not have to store the execution count of each basic block, only the total work.

Graph theory shows that we can calculate precisely the frequencies of all edges between basic blocks if we choose a spanning tree of the
function CFG and instrument the edges not in the tree. To make the instrumentation probe placement optimal, we choose the maximum spanning
tree, with the edge weights representing edge frequency estimates. As many high frequency edges as possible will be in the maximum spanning
tree and they will not be instrumented.

@@

\paragraph{Populating the edge flows:}
Intuitively, if all the edge flows are known for the complement of a spanning tree
then at any leaf of the spanning tree there is only one unknown edge flow.
This unknown edge flow can be calculated by Kirchhoff's first law.
This process repeats as a bottom-up propagation until all the unknown edge flows
have been calculated.
This algorithm can be formally defined as a post-order traversal on the spanning
tree.
Let $G$ be the CFG.
This algorithm first initializes the edge flows $D_{(u,v)}$, for each edge
$(u,v)$ in the CFG, as follows:
\[
D_{(u,v)} \gets
\begin{cases}
    P_{(u,v)} & \quad \text{if edge $(u,v)$ has a probe $P_{(u,v)}$}\\
    0       & \quad \text{otherwise}
\end{cases}
\]
Afterwards, for each vertex $u\in G$, in a post-order traversal of the spanning tree,
we can use the Kirchhoff's first law and apply the following operations
in a symbolic fashion:
\[
\Sigma^+_u = \sum_{v\in N^+(u)} D_{(u,v)}
\]
\[
\Sigma^-_u = \sum_{v\in N^-(u)} D_{(v,u)}
\]
\[
\forall v\in N^+(u):  D_{(u,v)} \gets
\begin{cases}
    D_{(u,v)} & \quad \text{if $D_{(u,v)}\neq 0$}\\
    \Sigma^-_u - \Sigma^+_u       & \quad \text{otherwise}
\end{cases}
\]
\[
\forall v\in N^-(u):  D_{(v,u)} \gets
\begin{cases}
    D_{(v,u)} & \quad \text{if $D_{(v,u)}\neq 0$}\\
    \Sigma^+_u - \Sigma^-_u       & \quad \text{otherwise}
\end{cases}
\]

@@

In this section we go one step even further.
In some cases, even the proposed relaxation strategy can still be conservative.
This conservatism can be overly restrictive in some cases, resulting in a
negligible overhead reduction but also causing just a negligible dynamic error
to the work profiling.
For these cases, we propose an adapted version of the relaxation algorithm that
operates on the whole program.
Traditionally, compiler optimizations are performed on the function-level, or at
best on a per module basis.
Whole program optimization (WPO) means that the compiler considers all
compilation units of the program and optimizes them using the combined knowledge
of how they are used together.

>>>>

In this section, we describe how to measure work online. Total work is just the sum of the costs of all executed basic blocks, so it is
possible to embed its computation into the execution of the program. A na\"{\i}ve instrumentation would use a global counter initialized at
the interception value, $\varepsilon$. Each basic block would then have only to increment this counter with the total cost of its
instructions every time it is executed. Although easy to implement, the na\"{\i}ve approach introduces a significant overhead.

Previous work on basic block profiling has proposed optimal ways for instrumenting the code with as little overhead as possible without
losing any profiling information~\citep{knuth73,ball94}. The underlying idea is to instrument only a subset of the basic blocks of a
function and use the function's control flow graph (CFG) to calculate execution counts for the rest. We can use the same idea for our
purposes. Work profiling and basic block profiling are similar, both need to count how many times each basic block was executed. The only
important difference is that we do not have to store the execution count of each basic block, only the total work.

Graph theory shows that we can calculate precisely the frequencies of all edges between basic blocks if we choose a spanning tree of the
function CFG and instrument the edges not in the tree. To make the instrumentation probe placement optimal, we choose the maximum spanning
tree, with the edge weights representing edge frequency estimates. By doing so, we leave as many high frequency edges as possible in the
maximum spanning tree and they will not be instrumented, allowing us to minimize the instrumentation overhead.

@@

\paragraph{Populating the edge flows:}
Intuitively, if all the edge flows are known for the complement of a spanning tree then at any leaf of the spanning tree there is only one
unknown edge flow. This unknown edge flow can be calculated by Kirchhoff's first law\FIXME{\cite{}}. This process repeats as a bottom-up
propagation until all the unknown edge flows have been calculated. This algorithm can be formally defined as a post-order traversal on the
spanning tree. Let $G$ be the CFG. This algorithm first initializes the edge flows $D_{(u,v)}$, for each edge $(u,v)$ in the CFG, as
follows:
\[
D_{(u,v)} \gets
\begin{cases}
    P_{(u,v)} & \quad \text{if edge $(u,v)$ has a probe $P_{(u,v)}$}\\
    0       & \quad \text{otherwise}
\end{cases}
\]
Afterwards, for each vertex $u\in G$, in a post-order traversal of the spanning tree,
we can use the Kirchhoff's first law and apply the following operations
in a symbolic fashion:
\[
\Sigma^+_u = \sum_{v\in N^+(u)} D_{(u,v)}
\]
\[
\Sigma^-_u = \sum_{v\in N^-(u)} D_{(v,u)}
\]
\[
\forall v\in N^+(u):  D_{(u,v)} \gets
\begin{cases}
    D_{(u,v)} & \quad \text{if $D_{(u,v)}\neq 0$}\\
    \Sigma^-_u - \Sigma^+_u       & \quad \text{otherwise}
\end{cases}
\]
\[
\forall v\in N^-(u):  D_{(v,u)} \gets
\begin{cases}
    D_{(v,u)} & \quad \text{if $D_{(v,u)}\neq 0$}\\
    \Sigma^+_u - \Sigma^-_u       & \quad \text{otherwise}
\end{cases}
\]

@@

In this section we go one step further. In some cases, even the proposed relaxation strategy can still be conservative. This conservatism
can be overly restrictive in some cases, resulting in a negligible overhead reduction but also causing just a negligible dynamic error to
the work profiling. For these cases, we propose an adapted version of the relaxation algorithm that operates on the whole program.
Traditionally, compiler optimizations are performed on the function-level, or at best on a per module basis. Whole program optimization
(WPO) means that the compiler considers all compilation units of the program and optimizes them using the combined knowledge of how they
are used together.
